P4: Explore and Summarize Data by Julia Kudinovich
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load the packages needed for analysis

library(ggplot2)
library(knitr)
library(dplyr)
library(gridExtra)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
wine <-read.csv('wineQualityWhites.csv')
```

# Intro to the dataset
This report explores a tidy dataset containing 4,898 white wines with 11 
variables on quantifying the chemical properties of each wine. At least 3 wine 
experts rated the quality of each wine, providing a rating between 0 (very bad) 
and 10 (very excellent).

In the report I will be investigating which chemical properties influence the 
quality of white wines.


# Univariate Analysis

## Statistics on the dataset

First, let's run some statistics on the dataset. 
```{r echo=FALSE, Univariate_Plots}
str(wine)
```

### What is the structure of your dataset?
The dataset contains 4898 observation of 13 variables: 12 features 
(fixed.acidity, volatile.acidity, citric.acid, residual.sugar, chlorides, 
free.sulfur.dioxide, total.sulfur.dioxide, density, pH, sulphates, 
alcohol and quality) and 1 id column

All variables are numeric. Quality is categorical discrete variable 
(rating form 0 to 10), all the other variables are continuous.

Below is the summary for each variable in the dataset:

```{r echo=FALSE}
summary(wine)
```

75% of wines have rating 6 or below with 3 being the lowest rating and 9 - 
highest.

### Did you create any new variables from existing variables in the dataset?

Quality is an integer field. It can only have values between 0 and 10. 
For better representation I would like to create factor variable out of quality.

```{r echo=FALSE}
wine$quality.factor<- ordered(wine$quality)
summary(wine$quality.factor)
```

Futhermore, I want to group quality factor and create 3 rating groups 
out of it. Wines with rating 3-5 will have 'low' factor, 6 'average' and
7-9 will be 'high' qualiuty factor. 

```{r echo=FALSE}
wine$quality.factor<- ifelse(wine$quality<=5, 'low', 'average')
wine$quality.factor<- ifelse(wine$quality>=7, 'high', wine$quality.factor)
wine$quality.factor <-ordered(wine$quality.factor, levels = c("low", "average", "high"))
summary(wine$quality.factor)

```


## What is/are the main feature(s) of interest in your dataset?

The main feature in the dataset is `quality` which determines how good or bad 
the wine is. I would like to investigate which chemical features influence the
main feature

### Quality

Let's graph quality:

```{r echo=FALSE}
qplot(x=quality, data=wine,  geom="bar") +
  scale_x_continuous(breaks=seq(3,9,1))
```

From the graph above we can see that the quality of wines has a bell shaped
normal distribution with most of the wines having 5 or 6 rating, which is 
consistent with 5.878 mean for wine quality we got from the summary. Very low 
number of wines have rating of 9 and none have very excellent (10) rating. 
Similarly, the lowest rating is 3 with very little number of wines having 
this rating.


## What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

I will investigate the chemical properties that affect the taste or smell
since the wine rating was given based on sensory data. Namely, features
I am interested are: `citric acid` (adds freshness and flavor to wines),
`residual sugar` (determines how sweet wine is), `alcohol` (can alter the taste),
`pH` (describes how acidic wine is).

### Alcohol

Summary:
```{r echo=FALSE}
summary(wine$alcohol)
```

Let's plot alcohol content:

```{r echo=FALSE}
qplot(x=alcohol, data=wine, binwidth=0.1) + 
  scale_x_continuous(breaks=seq(8,15,0.5))
```

From the above graph we can see that alcohol content distribution is positively 
skewed. More wines have lower alcohol.

Median alcohol content is 10.4 with most of the wines having alcohol content 
between 8 and 11.40


### Citric Acid

Summary:

```{r echo=FALSE}
summary(wine$citric.acid)
```

Plot:

```{r echo=FALSE}
qplot(x=citric.acid, data=wine, binwidth=0.01)+
  scale_x_continuous(breaks=seq(0,2,0.1))
```
 
 Most wines have citric acid between 0 and 0.39. 3rd quantile of 0.39 differs 
 a lot from the maximum value of 1.66. We can see that there are outliers with 
 values 0.7 and above. Also, there are a peaks around 0.5 and 0.75 values.


### Residual sugar

Summary:

```{r echo=FALSE}
summary(wine$residual.sugar)
```

Plot:

```{r echo=FALSE}
qplot(x=residual.sugar, data=wine, binwidth=0.5) 
```

Residual sugar is positively skewed and has a long right tail and several 
vary large valued outliers. 3rd quantile ha the value of 9.9 while the
maximum value for the variable is 65.8


### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

I found that residual.sugar distribution is positively skewed and has a long 
right tail. I am going to transform long tailed distribution using 
log scale



```{r echo=FALSE}
qplot(x=log(residual.sugar), data=wine, binwidth=0.1)+
  scale_x_continuous(breaks=seq(0,4,0.2))
 
```

After transformation we can see that distribution is bimodal with peaks around 
0.2 and 2.


### pH

Summary:

```{r}
summary(wine$pH)
```


Plot:

```{r echo=FALSE}
qplot(x=pH, data=wine, binwidth=0.05) +
  scale_x_continuous(breaks=seq(2,4, 0.1))

```

pH has a bell shaped normal distribution with a peak around 3.1.
Since pH is a logarithmic scale the wine with maximum value of 3.82 pH is ten
times more acidic than wine with minimum value of 2.72.


Below are several more graphs exploring remaining variables in the dataset.

```{r echo=FALSE}
p1 <- qplot(x=fixed.acidity, data=wine, binwidth=0.1) 
p2 <- qplot(x=volatile.acidity, data=wine, binwidth=0.01) 
p3 <- qplot(x=chlorides, data=wine, binwidth=0.005) 
p4 <- qplot(x=free.sulfur.dioxide, data=wine, binwidth=1) 
p5 <- qplot(x=total.sulfur.dioxide, data=wine, binwidth=1) 
p6 <- qplot(x=density, data=wine,  binwidth=0.001) 
p7 <- qplot(x=sulphates, data=wine, binwidth=0.01) 

grid.arrange(p1, p2, p3, p4, p5, p6, p7, ncol=2)
 
```

At the first glance, most of the above variables have a bell shaped distribution 
with some having right tails.


# Bivariate Plots Section

## Quality vs alcohol

First, I am interested in relationship between quality and alcohol.
Let's plot them below:

```{r echo=FALSE, Bivariate_Plots}
ggplot(aes(x=quality.factor, y=alcohol),data=wine) +
  geom_boxplot(alpha = 0.5,color = 'blue')+
  geom_jitter( alpha = 0.1)  +
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 4)
```

From the plot we can see apparent correlation between alcohol and quality. 
Better wines have higher alcohol.

Let's check correlation coefficient between alcohol and quality:

```{r echo=FALSE}
cor.test(wine$alcohol, wine$quality)
```

Correlation coefficient of 0.4355747 means that there is a positive meaningful 
correlation between alcohol content and quality.

## Quality vs residual sugar

Next, onto relationship between residual sugar and quality:

```{r echo=FALSE}
ggplot(aes(x=quality.factor, y=log(residual.sugar)),data=wine) +
  geom_boxplot(alpha = 0.5,color = 'blue')+
  geom_jitter( alpha = 0.1)  +
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 4)
```

```{r echo=FALSE}
cor.test(wine$residual.sugar, wine$quality)
```

Even though from the graph we may think there is a negative relationship 
between residual sugar and quality, the correlations coefficient is very low
which tells me that these 2 variables little if any correlation.



## Quality vs Citric acid 

```{r echo=FALSE}
ggplot(aes(x=quality.factor, y=citric.acid),data=wine) +
  geom_boxplot(alpha = 0.5,color = 'blue')+
  coord_cartesian(ylim=c(0,1.25)) +
  geom_jitter( alpha = 0.1)  +
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 4)
```
```{r echo=FALSE}
cor.test(wine$citric.acid, wine$quality)
```

Form the graph and correlation coefficient calculation we can conclude that
there is little if any correlation between quality and citric acid.


## Quality vs pH

```{r echo=FALSE}
ggplot(aes(x=quality.factor, y=pH),data=wine) +
  geom_boxplot(alpha = 0.5,color = 'blue')+
  geom_jitter( alpha = 0.1)  +
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 4)
```


```{r echo=FALSE}
cor.test(wine$pH, wine$quality)
```

Form the graph and correlation coefficient calculation we can conclude that
there is little if any correlation between quality and pH.


## pH vs acidity


```{r echo=FALSE, message=FALSE, warning=FALSE}
p1 <- ggplot(aes(x=pH, y=fixed.acidity),data=wine) +
  geom_jitter(alpha=0.1)+
  geom_smooth()
p2<- ggplot(aes(x=pH, y=volatile.acidity),data=wine) +
  geom_jitter(alpha=0.1)+
  geom_smooth()

grid.arrange(p1,p2, ncol=2)
```

```{r}
cor.test(wine$pH, wine$fixed.acidity)
cor.test(wine$pH, wine$volatile.acidity)
```

pH and fixed acidity have moderate correlation. The lower the fixed acidity 
the higher is pH. Which is not surprising since pH describes how acidic the
wine is on a scale from 0 (very acidic) to 14 (very basic).
On the other hand volatile acidity and pH have little to no correlation.


## Dencity vs Alcohol and Residual sugar

```{r echo=FALSE, message=FALSE, warning=FALSE}
p1 <- ggplot(aes(x=residual.sugar, y=density),data=wine) +
  geom_jitter(alpha=0.1)+
  coord_cartesian(xlim=c(0,30), ylim=c(0.9871,1.01)) +
  geom_smooth()
p2 <- ggplot(aes(x=alcohol, y=density),data=wine) +
  geom_jitter(alpha=0.1)+
  coord_cartesian(ylim=c(0.9871,1.01)) +
  geom_smooth()

grid.arrange(p1,p2, ncol=2)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
cor.test(wine$density, wine$residual.sugar)
cor.test(wine$density, wine$alcohol)

```


There is a strong positive almost linear relationship between residual sugar and 
density and negative relationship between alcohol and density with a couple of 
outliers. Correlation coefficients in both cases are quite large.


## Remaining variables

Next, I want to calculate correlation coefficients between quality and 
remaining variables to see the possibility of relationships I could have missed.


```{r echo=FALSE, message=FALSE, warning=FALSE}
cor.test(wine$quality, wine$fixed.acidity)
cor.test(wine$quality, wine$chlorides)
cor.test(wine$quality, wine$free.sulfur.dioxide)
cor.test(wine$quality, wine$total.sulfur.dioxide)
cor.test(wine$quality, wine$density)
cor.test(wine$quality, wine$volatile.acidity)
cor.test(wine$quality, wine$sulphates)
```

Out of remaining variables volatile acidity, density, total sulfur dioxide and
chlorides have small negative correlation with quality. Density has the most
significant coefficient among these variables.


I will graph variables mentioned above:


```{r echo=FALSE, message=FALSE, warning=FALSE}
p1 <-ggplot(aes(x=quality.factor, y=density),data=wine) +
  geom_boxplot()+
  coord_cartesian(ylim=c(0.9871, 1.005))
p2 <-ggplot(aes(x=quality.factor, y=volatile.acidity),data=wine) +
  geom_boxplot()+
  coord_cartesian(ylim=c(0, 0.6))
p3<-ggplot(aes(x=quality.factor, y=total.sulfur.dioxide),data=wine) +
  geom_boxplot()+
  coord_cartesian(ylim=c(0,300))
p4<-ggplot(aes(x=quality.factor, y=chlorides),data=wine) +
  geom_boxplot()+
  coord_cartesian(ylim=c(0,0.1))

grid.arrange(p1,p2,p3,p4, ncol=2)

```

From above we see that all the variables have negative correlation with quality.
The lower the variable value the higher the quality.


# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

I discovered some relationships between quality and several chemical features.

For example, there is a correlation between alcohol and quality. Better wines 
have higher alcohol content.

Density and quality have some correlation: the lower the density higher 
the quality.

Also, quality have negative relationships with volatile acidity, total sulfur 
dioxide and chlorides.


### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

I discovered that residual sugar and alcohol have strong relationships with 
density. 


### What was the strongest relationship you found?

The strongest relationship was density with residual sugar. Correlation 
coefficient for the relationship is 0.8389665 which means high positive 
correlation: the higher residual sugar is the higher density is.



# Multivariate Plots Section

## Volatile acidity vs total sulfur dioxide by quality

First I want to examine relationship between volatile acidity and total sulfur
dioxide with quality.


```{r echo=FALSE, Multivariate_Plots}
ggplot(aes(y=volatile.acidity, x=total.sulfur.dioxide, color=quality.factor),data=wine) +
  geom_jitter()+
  geom_smooth(method = "lm", se = FALSE,size=1)  +
  scale_colour_brewer( type = "seq", 
                       guide=guide_legend(title='Quality'),  
                       palette ='Set1')
```

It looks like better quality wines tend to concentrate closer the lower left 
corner than lower quality wines
This means that better wines have lower volatile acidity and total sulfur 
dioxide.


### Density vs alcohol vs residual sugar

In bivariate section I discovered that both alcohol and residual sugar have 
strong relationships with density. Now I want to see how these 3 variables
interact together.


```{r echo=FALSE}
ggplot(aes(x=density, y=residual.sugar, color=alcohol), data=wine) +
  facet_wrap(~quality.factor) +
  coord_cartesian(ylim = c(0, 25), xlim=c(0.9871, 1.005)) +
  geom_jitter() 

```

From the graph we see that wines with lower residual sugar and higher alcohol
are less dense than sweeter wines with less alcohol. What is interesting is 
that relationship between these 3 variables do not depend on quality of wines.
For different levels of quality, we see almost identical plots.



## Quality vs chlorides vs alcohol

```{r echo=FALSE}
ggplot(aes(x=chlorides, y=alcohol, color=quality.factor), data=wine) +
  geom_jitter() +
  coord_cartesian(xlim=c(0,0.15), ylim=c(8,14))+
  geom_smooth(method = "lm", se = FALSE,size=1)  +
  scale_colour_brewer( type = "seq", 
                       guide=guide_legend(title='Quality'),  
                       palette ='Set1')

```

Quality seem to increase with increasing alcohol and descreasing chlorides



# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

It became clear that low volatile acidity, low total sulfur dioxide, high 
alcohol and low chlorides contribute to the better wine quality


### Were there any interesting or surprising interactions between features?

What I found interesting is that relationship between density, alcohol and 
residual sugar is independent of wine quality.


------

# Final Plots and Summary

### Plot One: Alcohol vs Quality Plot
```{r echo=FALSE, Plot_One}
ggplot(aes(x=quality.factor, y=alcohol, fill=quality.factor),data=wine) +
   geom_jitter( alpha = .3)  +
  geom_boxplot( alpha = .5,color = 'blue')+
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 4) +
  ggtitle("Alcohol level by wine quality")+
  ylab('Alcohol Content (%)') +
  xlab('Quality') 

```

### Description

I chose this plot because it represents the strongest relationship out of all 
I have found. 
It shows that with increasing alcohol content quality of wine goes up


### Plot Two: Density vs alcohol and residual sugar
```{r echo=FALSE, Plot_Two}
p1 <- ggplot(aes(x=residual.sugar, y=density, color=quality.factor),data=wine) +
  geom_jitter(alpha=0.1)+
  coord_cartesian(ylim = c(0.9871, 1.005), xlim=c(0,30)) +
  geom_smooth(method = "lm", se = FALSE,size=1)  +
  ggtitle("Density vs Residual Sugar")+
  ylab('Density (gram/liter)') +
  xlab('Residual sugar (gram/liter)')+
  scale_colour_brewer( type = "seq", 
                       guide=guide_legend(title='Quality'),  
                       palette ='Set1')

p2 <- ggplot(aes(x=alcohol, y=density, color=quality.factor),data=wine) +
  geom_jitter(alpha=0.1)+
  coord_cartesian(ylim = c(0.9871, 1.005)) +
  geom_smooth(method = "lm", se = FALSE,size=1)  +
  ggtitle("Density vs Alcohol")+
  ylab('Density (gram/liter)') +
  xlab('Alcohol content (%)')+
  scale_colour_brewer( type = "seq", 
                       guide=guide_legend(title='Quality'),  
                       palette ='Set1')


grid.arrange(p1,p2, ncol=2)
```

```{r echo=FALSE}

ggplot(aes(x=density, y=residual.sugar, color=alcohol), data=wine) +
  geom_jitter()+
  coord_cartesian(ylim = c(0, 25), xlim=c(0.9871, 1.005)) +
  ggtitle("Residual sugar vs Density by Alcohol")+
  ylab('Residual sugar (gram/liter)') +
  xlab('Density (gram/liter)')+
  facet_wrap(~quality.factor)
```


### Description Two

I found 2 very strong relationships between density and alcohol (negative) and
between density and residual sugar (positive). Both these relationships have 
high close to linear correlation.

What is more interesting, when put all these 3 variables against each other we
see that relationship between these 3 variables do not depend on quality of 
wines. For different levels of quality, we see almost identical plots.



### Plot Three: Volatile acidity vs total sulfur dioxide by quality
```{r echo=FALSE, Plot_Three}
ggplot(aes(y=volatile.acidity, x=total.sulfur.dioxide, color=quality.factor),data=wine) +
  geom_jitter(alpha=0.1)+
  coord_cartesian(xlim=c(0, 350), ylim=c(0,1)) +
  geom_smooth(method = "lm", se = FALSE,size=1) + 
  ggtitle("Volatile acidity vs total sulfur dioxide by quality")+
  ylab('Volatile acidity (gram/liter)') +
  xlab('Total sulfur dioxide (parts per million)')+
  scale_colour_brewer( type = "seq", 
                       guide=guide_legend(title='Quality'),  
                       palette ='Set1')
```

### Description Three
Plot shows relationship between volatile acidity and total sulfur dioxide for
different wine quality. Better wines have lower volatile acidity and total 
sulfur dioxide.


------

# Reflection

It is common to think that wine tastings are subjective to the individual.
However, in the analysis I had success in finding relationships between wine 
quality and several chemical factors. 

The main factors that influence wine quality are alcohol (strong, positive
relationship), density. volatile acidity, total sulfur dioxide and chlorides
(all having negative relationship).

What surprised me is that I have not found a relationship between quality
and citric acid which is supposed to add freshness and taste to wines. I 
believe further analysis should be done on citric acid to possibly uncover
the relationship.

Also, I did not find a relationship between pH level and quality. That 
particular variable was among the ones I thought would have a great impact on 
quality. 

One of the struggles I came across was the choice of variables that might 
influence wine quality. Without having a chemistry and/or wine background 
knowledge I found it hard to pick variables.  From the description of chemical 
factors and common sense I have selected several that I thought might influence 
quality, However, only after I explored each variable one at a time I found out
which of those variables actually had relationships.

Also, I need to mention that correlation does not mean causation. All of the
findings suggest that there might be a relationship however we cannot say
that particular chemical factor caused better wine quality. For the possible
future analyses it would be beneficial to carry a controlled expertise to
identify if there is a causation.


